---
title: "MMA & Lung Cancer Analysis"
author: "Rachel Heise, MS"
pi: "Yiwey Shieh, MD"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  wcmtheme::wcm_html: 
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(survival)
library(tidyverse)
library(gtsummary)

load("/Volumes/Shieh-share$/MMA_Cancer/output/matched_data_3.RData")
# load("/Volumes/Shieh-share$/MMA_Cancer/output/matched_data_4.RData")
# load("/Volumes/Shieh-share$/MMA_Cancer/output/matched_data_cal2.RData")
# load("/Volumes/Shieh-share$/MMA_Cancer/output/matched_data_3_adj.RData")
# load("/Volumes/Shieh-share$/MMA_Cancer/output/matched_data_3_cat.RData")

matched_data_3 <- matched_data_3 %>% 
  mutate(MMA_quartile = factor(ntile(MMA, n = 4)))

```

## Conditional Logistic Regression - 1:3 Optimal Matching

This is a 1:3 matched case-control study, assessing if there is an association between MMA (nmol/L) and lung cancer. 24 cases (individuals with lung cancer, pre-treatment) were each matched with 3 controls from the NHANES dataset for a total of 96 individuals included in the analysis. Matching variables were age, B12 (pg/mL), and Creatinine (mg/dL). 
  
The below model is a conditional logistic regression (CLR) model which takes matching into account. The outcome in the CLR model is lung cancer. MMA, B12, and Creatinine are all highly skewed covariates so the log of each is included in the model. The coefficient of log(MMA) has been standardized such that the odds ratio for that covariate represents the odds of lung cancer for a one standard deviation increase in log(MMA).

```{r}
# take the log of MMA and standardize it to calculate the standardized OR

matched_data_3 <- matched_data_3 %>% 
  mutate(log_MMA = log(MMA),
         log_MMA_standardized = log(MMA)/sd(log(MMA)))

# matched_data_3 %>% 
#   ggplot(aes(x = log(B12))) +
#   geom_histogram(bins = 15)

model <- clogit(cancer ~ log_MMA_standardized + age + log(B12) + 
                  log(CREATININE) + strata(subclass),
                data = matched_data_3, id = subclass)

model %>% 
  tbl_regression(exp = T,
                 estimate_fun = purrr::partial(style_ratio, digits = 4)) %>%
  bold_p()
```


Holding age, B12, and Creatinine constant, for a one standard deviation increase in log MMA, the odds of having lung cancer are 130% higher. The odds ratio (OR) is 2.30 (95% CI: 1.29, 4.08).


### MMA quartiles

Using the same matched data, the below model uses a CLR with MMA in terms of quartiles rather than continuous. Quartiles were calculated using all individuals (both cases and controls), and quartile 1 represents the lowest values of MMA and quartile 4 represents the highest values of MMA.

```{r}
model2 <- clogit(cancer ~ MMA_quartile + age + B12 + CREATININE + strata(subclass),
                 data = matched_data_3, id = subclass)

model2 %>%
  tbl_regression(exp = T,
                 estimate_fun = purrr::partial(style_ratio, digits = 2),
                 include = MMA_quartile) %>%
  bold_p()
```

Holding age, B12, and Creatinine constant, for someone in the highest quartile of MMA, the odds of having lung cancer are 5.69 times that of someone in the lowest quartile. The odds ratio (OR) is 5.69 (95% CI: 1.17, 27.73).


<br>



```{r, message=FALSE, echo = FALSE, eval=FALSE}
## Check model assumptions

model_untransformed <- glm(cancer ~ MMA + B12 + CREATININE + age,
                           data = matched_data_3, family = binomial)

probabilities <- predict(model_untransformed, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
head(predicted.classes)

predictors <- c("MMA", "B12", "CREATININE", "age")
data_untransformed <- matched_data_3 %>%
  select(MMA, B12, CREATININE, age) %>% 
  mutate(logit = log(probabilities/(1-probabilities))) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

ggplot(data_untransformed, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y") +
  labs(title = "Untransformed model")
```

```{r, message=FALSE, echo = FALSE, eval=FALSE}
data_transformed <- matched_data_3 %>% 
  mutate(log_B12 = log(B12),
         log_CREATININE = log(CREATININE))

model_transformed <- glm(cancer ~ log_MMA + log_B12 + log_CREATININE + age,
                             data = data_transformed, family = binomial)

probabilities <- predict(model_transformed, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
# head(predicted.classes)

predictors <- c("log_MMA", "log_B12", "log_CREATININE", "age")
data_transformed <- data_transformed %>%
  select(log_MMA, log_B12, log_CREATININE, age) %>% 
  mutate(logit = log(probabilities/(1-probabilities))) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

ggplot(data_transformed, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y") +
  labs(title = "Log-transformed model")
```


```{r, eval=FALSE, include=FALSE}
matched_data_3_cases <- matched_data_3 %>% 
  filter(cancer == 1)

matched_data_3_controls <- matched_data_3 %>% 
  filter(cancer == 0)

matched_data_3 %>% 
  ggplot(aes(x = CREATININE, fill = as.factor(cancer))) +
  geom_density(alpha = 0.5)

# try to avoid sparse data problem by filtering on CREATININE < 1.01 which is the max value of CREATININE
# in the cases group

# matched_data_3_creatinine <- matched_data_3 %>% 
#   filter(CREATININE <= 1.01)
# 
# model2 <- clogit(cancer ~ MMA + age + B12 + CREATININE + strata(subclass),
#                 data = matched_data_3_creatinine, id = subclass)
# summary(model2)
# 
# model2 %>% 
  # tbl_regression(exp = T,
  #                estimate_fun = purrr::partial(style_ratio, digits = 4)) %>%
  # bold_p()

```

```{r, eval=FALSE, include=FALSE}
# try analysis with 1:4 data

model4 <- clogit(cancer ~ MMA + age + B12 + CREATININE + strata(subclass),
                data = matched_data_4, id = subclass)
# summary(model)

model4 %>% 
  tbl_regression(exp = T,
                 estimate_fun = purrr::partial(style_ratio, digits = 4)) %>%
  bold_p()
```



