---
title: "NHANES MMA Data"
author: "Rachel Heise, MS"
pi: "Yiwey Shieh, MD"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  wcmtheme::wcm_html: 
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(gtsummary)
library(haven)
library(labelled)
library(rstatix) #identify outliers

set_gtsummary_theme(theme_gtsummary_compact())
theme_set(theme_minimal())

load(here::here("./output/NHANES_dat.RData"))

# remove patients without complete MMA, B12, and creatinine data
nhanes_dat <- nhanes_dat %>% 
  filter(complete_data == 1)
```

## Notes

* This data was pulled from NHANES (National Health and Nutrition Examination Survey)
* MMA data was available for the survey years 2001-2002, 2003-2004, 2011-2012, and 2013-2014
* Data for participants who have complete MMA, B12, and serum creatinine data are included in this report


## Demographics

```{r}
nhanes_dat %>% 
  select(gender, age, age_cat, race_ethnicity, year, B12, CREATININE, MMA,
         cancer_malignancy, lung_cancer) %>% 
  mutate(lung_cancer = if_else(is.na(lung_cancer), 0, lung_cancer),
         cancer_malignancy = if_else(is.na(cancer_malignancy), "No", cancer_malignancy)) %>% 
  tbl_summary(label = list(age ~ "Age",
                           age_cat ~ "Age (categorical)",
                           year ~ "NHANES year",
                           race_ethnicity ~ "Race/ethnicity",
                           lung_cancer ~ "Ever had lung cancer",
                           cancer_malignancy ~ "Ever had cancer or malignancy")) %>% 
  bold_labels()
```

## Lab results by age category

### MMA

```{r, fig.width=5, fig.height=4}
nhanes_dat %>% 
  ggplot(aes(x = age_cat, y = log(MMA))) +
  geom_boxplot() +
  labs(x = "Age category",
       title = "log(MMA) by Age Category")
```

<br>

```{r, fig.width=5, fig.height=4}
nhanes_dat %>% 
  filter(B12 > 300, CREATININE < 1.3) %>% 
  ggplot(aes(x = age_cat, y = MMA)) +
  geom_boxplot() +
  labs(x = "Age category",
       title = "MMA for participants with B12 > 300 and \nserum creatinine < 1.3")
```

<br>

```{r, fig.width=5, fig.height=4}
nhanes_dat %>% 
  filter(B12 > 300, CREATININE < 1.3) %>% 
  ggplot(aes(x = age_cat, y = log(MMA))) +
  geom_boxplot() +
  labs(x = "Age category",
       title = "log(MMA) for participants with B12 > 300 and \nserum creatinine < 1.3")
```

<br>

```{r, fig.width=5, fig.height=4, message=FALSE}
nhanes_dat %>% 
  filter(B12 > 300, CREATININE < 1.3) %>%
  filter(MMA <1000) %>%
  ggplot(aes(x = MMA)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 410) +
  labs(title = "MMA for participants with B12 > 300 and serum \ncreatinine < 1.3",
       subtitle = "Note: MMA is restricted to values < 1000 for better visual \nassessment",
       y = "count") +
  annotate("text", x = 570, y = 1000, label ="Extreme Outliers")
```

<br>

### Table of MMA values by Age

```{r}
# Summarize median and IQR for each level of the categorical variable
summary_table_all <- nhanes_dat %>%
  group_by(age_cat) %>%
  summarise(`All participants` = paste(
      sprintf("%.0f", median(MMA)),
      sprintf("(%.0f, %.0f)", quantile(MMA, 0.25), quantile(MMA, 0.75))))


summary_table_restricted <- nhanes_dat %>%
  filter(B12 > 300, CREATININE < 1.3) %>% 
  group_by(age_cat) %>%
  summarise(`Only participants with B12 > 300 and creatinine < 1.3` = paste(
      sprintf("%.0f", median(MMA)),
      sprintf("(%.0f, %.0f)", quantile(MMA, 0.25), quantile(MMA, 0.75))))

summary_table <- summary_table_all %>% 
  full_join(summary_table_restricted, by = "age_cat") %>% 
  rename(`Age category` = age_cat)

# Display the summary table
summary_table %>% 
  kableExtra::kable() %>% 
  kableExtra::kable_styling(full_width = FALSE) %>% 
  kableExtra::add_header_above(c("Median (Q1, Q3) of MMA" = 3)) %>% 
  kableExtra::column_spec(3, width = "12em")
```


<br>

### B12

```{r, fig.width=5, fig.height=4}
nhanes_dat %>% 
  ggplot(aes(x = age_cat, y = log(B12))) +
  geom_boxplot() +
  labs(x = "Age category",
       y = "log(B12) (pg/mL)")
```


<br>

### Creatinine
<br>

```{r, fig.width=5, fig.height=4}
nhanes_dat %>% 
  ggplot(aes(x = age_cat, y = log(CREATININE))) +
  geom_boxplot() +
  labs(x = "Age category", y = "log(Serum Creatinine) (mg/dL)")
```

```{r, eval=FALSE}
# try continuous age (scatterplot)
nhanes_dat %>% 
  ggplot(aes(x = age, y = CREATININE)) +
  geom_point()
```


## Extreme Outliers

Extreme outliers are defined as values above $Q3 + 3 \cdot IQR$, or below $Q1 - 3 \cdot IQR$.

```{r, message = FALSE}
mma_outliers <- identify_outliers(nhanes_dat, variable = "MMA") %>% 
  filter(is.extreme == TRUE)
creatinine_outliers <- identify_outliers(nhanes_dat, variable = "CREATININE") %>% 
  filter(is.extreme == TRUE)
b12_outliers <- identify_outliers(nhanes_dat, variable = "B12") %>% 
  filter(is.extreme == TRUE)

nhanes_dat %>% 
  mutate(`MMA extreme outlier` = if_else(SEQN %in% mma_outliers$SEQN, 1, 0),
         `B12 extreme outlier` = if_else(SEQN %in% b12_outliers$SEQN, 1, 0),
         `Creatinine extreme outlier` = if_else(SEQN %in% creatinine_outliers$SEQN, 1, 0)) %>% 
         # extreme_outlier_and_cancer = if_else((`MMA extreme outlier` == 1 | 
         #                                        `B12 extreme outlier` == 1 |
         #                                        `Creatinine extreme outlier` == 1) & (cancer_malignancy == "Yes"), 1, 0),
         # total_extreme_outliers = `MMA extreme outlier`+ `B12 extreme outlier` + `Creatinine extreme outlier`) %>% 
  select(contains("outlier")) %>% 
  tbl_summary() %>% 
  bold_labels()
```

## Cancer

```{r}
cancer_pt <- nhanes_dat %>% 
  mutate(MMA_extreme_outlier = if_else(SEQN %in% mma_outliers$SEQN, 1, 0),
         B12_extreme_outlier = if_else(SEQN %in% b12_outliers$SEQN, 1, 0),
         Creatinine_extreme_outlier = if_else(SEQN %in% creatinine_outliers$SEQN, 1, 0)) %>%
  filter(cancer_malignancy == "Yes")
```

Of the `r nrow(cancer_pt)` participants who have had a cancer diagnosis, `r sum(cancer_pt$MMA_extreme_outlier == 1 | cancer_pt$B12_extreme_outlier == 1 | cancer_pt$Creatinine_extreme_outlier == 1)` participants have at least one extreme outlier (MMA, B12, or Creatinine). `r sum(cancer_pt$lung_cancer, na.rm = T)` participants have had a lung cancer diagnosis.

### Lung Cancer

```{r}
nhanes_dat %>% 
  filter(!is.na(age_lung_cancer_diag)) %>% 
  ggplot(aes(x = age, y = age_lung_cancer_diag)) +
  geom_point() +
  labs(x = "Age at time of NHANES study", 
       y = "Age at lung cancer diagnosis")
```

